---
title: "Week 1 Notes"
author: "Enes Kemal Ergin"
date: "Tuesday, February 17, 2015"
output: html_document
---

## Getting started with R

### Installing R

The first step is to install R. There are several resources on the internet on how to do this. But here are two:

* [Installing R on Windows](http://youtu.be/mfGFv-iB724)
* [Installing R on Mac](http://youtu.be/Icawuhf0Yqo)
* [Installing R on Ubuntu](http://cran.r-project.org/bin/linux/ubuntu/README)

### Installing RStudio

The next step is to install RStudio. Technically you can run all the code shown here without installing RStudio, but we highly recommend this integrated development environment (IDE). Instructions are [here](http://www.rstudio.com/products/rstudio/download/)

### Installing packages

The first R command we will run is `install.packages`. R only includes a basic set of functions. There is much more it can do than this, but not everybody needs everything so we instead make some functions via packages. Many of these function are stored in the Comprehensive R Archive Network (CRAN). Note that these packages are vetted. You can install easily from within R if you know the name of the function. We are going to install the package `devtools` 

```{r,eval=FALSE}
install.packages("devtools")
```

In RStudio you can go click on "Tools" and then "Install Packages"

Note to Windows users: To use devtools you will have to also install `Rtools`. In general you will need to install packages as administrator. One way to do this is to start R as administrator. If you do not have permission to do this, then it is a bit [more complicated](http://www.magesblog.com/2012/04/installing-r-packages-without-admin.html). 

The reason we installed this package is that we can also install packages from github. These packages are not vetted but many experimental packages are on github because they are not yet ready for CRAN. An example, is the `dagdata` package that we use extensively here. We will load the library and use one of its functions.

```{r}
library(devtools)
install_github("genomicsclass/dagdata")
```

### Learn R basics 

By following along these notes you will learn R and Statistics simultaneously. However, we will not teach the basics of R syntax because there are many resources already available. 

One resource that we highly recommend is [swirl](http://swirlstats.com/). Once you have R installed you can install it and run it this way:

```{r, eval=FALSE}
install.packages("swirl")
library(swirl)
swirl()
```

If you have never programmed in R, we recommend going through the entire R Programming tutorial.

An other alternative is [TryR](http://tryr.codeschool.com/)

## Importing data into R

The first step when getting ready to analyze data is to read in the data into R. There are several ways to do this and we will discuss three of them. But you only need to learn one. 

In the life sciences, small datasets such as the one used as an example in the next sections are stored as Excel file. In general you want to avoid the Excel (xls) format and save files as comma delimited (csv) or tab delinted (txt) files. The first step is to find the file containing your data and know it's *path*. 

The three files we start with in the class are these:

* [female mouse weights](https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/femaleMiceWeights.csv)
* [female control population](https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/femaleControlsPopulation.csv)
* [entire population dataset](https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/mice_pheno.csv)

You can navigate to these file by visiting the data directory of dagadat on [github](https://github.com/genomicsclass/dagdata/tree/master/inst/extdata)
### The working directory

When you are working in R it is useful to know your _working directory_. This is the directory or folder in which R will save or look for files by default. You can see your working directory by typing:

```{r}
getwd()
```

You can also change your working directory using the function `setwd`. Or you can change it through RStudio by clicking on "Session". 

The functions that read and write files (there are several in R)
 assume you meant to look for files or write files in the working directory. Our recommended approach for beginners will have you  reading and writing to the working directory. But you can also type the [full path](http://www.computerhope.com/jargon/a/absopath.htm) which will work independently of the working directory.

### Projects in RStudio

We find that the simplest way to organize yourself is to start a Project in RStudio (Click on "File"" and the "New Project"). When creating the project, you will select a folder to be associated with it. You can then download all your data into this folder. 

For example, as explained [here](https://www.youtube.com/watch?v=812ruYN4PZQ) a file we are using can be found our github repository [here](https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/femaleMiceWeights.csv). Save it as a csv file to your project folder. Then, assuming you have opened your project, you should be able to read it in like this

```{r,eval=FALSE}
dat <- read.csv("femaleMiceWeights.csv")
```

### Reading directly from github

You can also use R to download the file and define the path yourself. In this example, we are using the `download.file` function to download the file to a specific location and then read it in. We assign it a random name and a random directory using the function `tempfile` but you can change this so the file is saved a directory of your choosing.

```{r,message=FALSE}
library(downloader) ##use install.packages to install
url <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/femaleMiceWeights.csv"
filename <- tempfile()
download(url,destfile=filename)
```

Now we are ready to read in the file 
```{r}
dat <- read.csv(filename)
```

### Reading data stored in a package
The files you are downloading are actually stored in the 'dagdata' package. If you have the package installed the file is already in your system and you can use the 'system.file' function to find it

```{r}
dir <- system.file(package="dagdata")
list.files(dir)
list.files(file.path(dir,"extdata"))
filename <- file.path(dir,"extdata/femaleMiceWeights.csv")
dat <- read.csv(filename)
```


Now that you know some of the basics how to load data into R we can get started.

## Data summaries: summary, str

First we load an example data frame:

```{r}
rats <- data.frame(id = paste0("rat",1:10),  
                   sex = factor(rep(c("female","male"),each=5)),
                   weight = c(2,4,1,11,18,12,7,12,19,20),
                   length = c(100,105,115,130,95,150,165,180,190,175))
rats
```

The `summary` and `str` functions are two helpful functions for getting a sense of data. `summary` works on vectors or matrix-like objects (including data.frames). `str` works on an arbitrary R object and will compactly display the structure.

```{r}
summary(rats)
summary(rats$weight)
str(rats)
```

## Aligning two objects: match, merge

We load another example data frame, with the original ID and another secretID. Suppose we want to sort the original data frame by the secretID.

```{r}
ratsTable <- data.frame(id = paste0("rat",c(6,9,7,3,5,1,10,4,8,2)),
                        secretID = 1:10)
ratsTable
# wrong!
cbind(rats, ratsTable)
```

`match` is a very useful function in R, which can give us this order, but it's easy to get its arguments mixed up. Remember that `match` gives you, for each element in the first vector, the index of the first match in the second vector. So typically the data.frame or vector you are reordering would appear as the second argument to `match`. It's always a good idea to check that you got it right, which you can do by using `cbind` to line up both data frames.

```{r}
match(ratsTable$id, rats$id)
rats[match(ratsTable$id, rats$id),] 
cbind(rats[match(ratsTable$id, rats$id),], ratsTable)
```

Or you can use the `merge` function which will handle everything for you. You can tell it the names of the columns to merge on, or it will look for columns with the same name.

```{r}
ratsMerged <- merge(rats, ratsTable, by.x="id", by.y="id")
ratsMerged[order(ratsMerged$secretID),]
```

## Analysis over groups: split, tapply, and dplyr libary

Suppose we need to calculate the average rat weight for each sex. We could start by splitting the weight vector into a list of weight vectors divided by sex. `split` is a useful function for breaking up a vector into groups defined by a second vector, typically a factor. We can then use the `lapply` function to calculate the average of each element of the list, which are vectors of weights.

```{r}
sp <- split(rats$weight, rats$sex)
sp
lapply(sp, mean)
```

A shortcut for this is to use `tapply` and give the function which should run on each element of the list as a third argument:

```{r}
tapply(rats$weight, rats$sex, mean)
```

R is constantly being developed in the form of add-on packages, which can sometimes greatly simplify basic analysis tasks. A new library "dplyr" can accomplish the same task as above, and can be extended to many other more complicated operations. The "d" in the name is for data.frame, and the "ply" is because the library attempts to simplify tasks typically used by the set of functions: `sapply`, `lapply`, `tapply`, etc. Here is the same task as before done with the dplyr functions `group_by` and `summarise`:

```{r}
library(dplyr)
sexes <- group_by(rats, sex)
summarise(sexes, ave=mean(weight))
```

With dplyr, you can chain operations using the `%.%` operator:

```{r}
rats %.% group_by(sex) %.% summarise(ave=mean(weight))
```
